# Geo-Spatial Analysis and Visualization

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(tidyverse)
library(sf)
library(leaflet)
library(tidycensus)
library(tigris)
options(tigris_use_cache = TRUE)
```

Geo-spatial data is the easiest to make a methodological mistake with — we'll be layering data points on top of a map, indicating location, and we need to make sure our visualizations successfully show the right data.

The most common error in geo-spatial analysis is not accounting for population. If I told you there were more gun deaths in California in 2020 than in Mississippi, I could show data like this:

```{r, table1, echo=FALSE}
State <- c("California", "Mississippi")
Firearm_Deaths <- c("3,449", "818")
df <- data.frame(State, Firearm_Deaths)
library(knitr)
knitr::kable(df)
```

(Data from the [CDC](https://www.cdc.gov/nchs/pressroom/sosmap/firearm_mortality/firearm.htm))

Does California have a bigger gun problem than Mississippi? That would be the incorrect conclusion. When comparing states with different populations, we have to calculate a *rate* — in this case, firearm deaths *per capita*. The CDC has already done this, and it shows California ranks 44th in firearm death rate while Mississippi ranks 1st. The point: always normalize for population when comparing geographic units.

A second common error is reversed or incorrectly signed latitude/longitude values. When I map crimes in Philadelphia, I often get a cluster of data points in *Florida*. Someone made a mistake when entering coordinates, and repeated it many times. Always sanity-check your map: if markers appear in the wrong state or ocean, check whether latitude and longitude have been swapped, or whether a negative sign is missing.

## Getting Started: tidycensus

In this chapter we'll build maps using two key packages:

- **tidycensus** — pulls data from the U.S. Census Bureau's American Community Survey (ACS) directly into R, with optional built-in geometry, eliminating the need to manually download shapefiles or GeoJSON files.
- **leaflet** — renders interactive maps in R that can be zoomed, panned, and clicked.

Install them if you haven't already:

```{r, install, eval=FALSE}
install.packages('tidycensus')
install.packages('leaflet')
install.packages('tigris')
install.packages('sf')
```

### Getting a Census API Key

`tidycensus` requires a free API key from the Census Bureau. Get one here:

<https://api.census.gov/data/key_signup.html>

Once you have it, set it for your session:

```{r, apikey, eval=FALSE}
library(tidycensus)
census_api_key("YOUR_KEY_HERE")
```

You only need to do this once per session. To save it permanently so you never have to set it again, add `install = TRUE`:

```{r, apikey2, eval=FALSE}
census_api_key("YOUR_KEY_HERE", install = TRUE)
```

### Finding the Right Variable

The ACS contains thousands of variables. Use `load_variables()` to search for what you need:

```{r, vars, eval=FALSE}
vars <- load_variables(2022, "acs5")
View(vars)
```

This opens a searchable table. For example, `B19013_001` is median household income — one of the most commonly mapped variables.

## Your First tidycensus Map

The core function is `get_acs()`. Here's what a basic call looks like:

```{r, get_acs_example, eval=FALSE}
location <- get_acs(
  state = "",       # state abbreviation, e.g. "NC"
  county = "",      # county name (optional)
  geography = "",   # "state", "county", or "tract"
  variables = "",   # ACS variable code
  geometry = TRUE,  # include map geometry
  year = 2022
)
```

Setting `geometry = TRUE` is the key difference from a regular data pull — it attaches the polygon shapes for each geographic unit, so you can map the results directly without downloading any external files.

Let's map median household income by census tract in Mecklenburg County, NC (Charlotte):

```{r, clt_income}
income <- get_acs(
  geography = "tract",
  variables = "B19013_001",
  state = "NC",
  county = "Mecklenburg",
  year = 2022,
  geometry = TRUE
)
```

We now have a data frame that includes the Census geometry for each tract, plus the income estimate. Before mapping, we set the coordinate projection to WGS84 — this is the standard that Leaflet uses:

```{r, transform}
income |>
  st_transform('+proj=longlat +datum=WGS84') -> income
```

Now define a color palette. `colorBin` groups a continuous variable (income) into a fixed number of color bins:

```{r, clt_palette}
pal_clt <- colorBin("Blues", domain = income$estimate, 5)
```

And map it:

```{r, clt_map}
leaflet() |>
  addTiles() |>
  addPolygons(
    data = income,
    color = "black",
    weight = 1,
    fillColor = ~pal_clt(income$estimate),
    fillOpacity = 0.7,
    label = ~paste0("$", income$estimate)
  )
```

Each polygon is a census tract, shaded by median household income. Hover over a tract to see the dollar amount.

### Changing the Geography

The same code works at any geographic level. To map all counties in a state instead of tracts in one county, remove the `county` argument and change `geography` to `"county"`:

```{r, nc_snap}
snap <- get_acs(
  geography = "county",
  variables = c(snap_pct = "S2201_C03_001E"),
  state = "NC",
  year = 2023,
  survey = "acs5",
  geometry = TRUE
) |>
  st_transform(4326)

pal_snap <- colorBin(
  "YlGnBu",
  domain = snap$estimate,
  na.color = "#cccccc"
)

leaflet(snap) |>
  addTiles() |>
  addPolygons(
    fillColor = ~pal_snap(estimate),
    weight = 0.5,
    color = "#555",
    opacity = 1,
    fillOpacity = 0.8,
    label = ~paste0(NAME, ": ", estimate, "%"),
    highlightOptions = highlightOptions(
      weight = 2,
      color = "#333",
      bringToFront = TRUE
    )
  )
```

This maps the percentage of households receiving SNAP benefits in each NC county. Notice we're using `S2201_C03_001E` from the ACS Subject Tables — a different table than the standard `B` tables. The `load_variables()` function is the best way to find variable codes.

## Comparing Data Across Time

One of the most powerful things you can do with `tidycensus` is pull multiple years at once using `map_dfr()`, which applies a function to each year and stacks the results into one data frame. Then we can use Leaflet's layer controls to let the reader switch between years.

Here we map per-capita income in Philadelphia by census tract from 2019 to 2023:

```{r, philly}
years_to_get <- 2019:2023

philly <- map_dfr(years_to_get, \(yr) {
  get_acs(
    geography = "tract",
    variables = c(pop = "B01003_001", pc_inc = "B19301_001"),
    state = "PA",
    county = "Philadelphia",
    survey = "acs5",
    geometry = TRUE,
    year = yr,
    output = "wide"
  ) |>
    transmute(
      GEOID, NAME, yr,
      geometry,
      pop = popE,
      income_per_capita = pc_incE
    )
})
```

Split the combined data frame into one per year, and define a palette across the full range of values so colors stay consistent when toggling between years:

```{r, philly_split}
income2019 <- philly |> filter(yr == 2019)
income2020 <- philly |> filter(yr == 2020)
income2021 <- philly |> filter(yr == 2021)
income2022 <- philly |> filter(yr == 2022)
income2023 <- philly |> filter(yr == 2023)

pal_philly <- colorNumeric(
  "Greens",
  domain = philly$income_per_capita
)
```

Then build a map with a layer control that lets the reader select a year:

```{r, philly_map}
leaflet() |>
  addTiles() |>
  addPolygons(
    data = income2019, group = "2019",
    weight = 1, fillOpacity = 0.8, color = "black",
    fillColor = ~pal_philly(income_per_capita),
    label = ~paste0("$", income_per_capita)
  ) |>
  addPolygons(
    data = income2020, group = "2020",
    weight = 1, fillOpacity = 0.8, color = "black",
    fillColor = ~pal_philly(income_per_capita),
    label = ~paste0("$", income_per_capita)
  ) |>
  addPolygons(
    data = income2021, group = "2021",
    weight = 1, fillOpacity = 0.8, color = "black",
    fillColor = ~pal_philly(income_per_capita),
    label = ~paste0("$", income_per_capita)
  ) |>
  addPolygons(
    data = income2022, group = "2022",
    weight = 1, fillOpacity = 0.8, color = "black",
    fillColor = ~pal_philly(income_per_capita),
    label = ~paste0("$", income_per_capita)
  ) |>
  addPolygons(
    data = income2023, group = "2023",
    weight = 1, fillOpacity = 0.8, color = "black",
    fillColor = ~pal_philly(income_per_capita),
    label = ~paste0("$", income_per_capita)
  ) |>
  addLayersControl(
    baseGroups = c("2019", "2020", "2021", "2022", "2023"),
    options = layersControlOptions(collapsed = FALSE)
  ) |>
  addLegend(
    pal = pal_philly,
    values = philly$income_per_capita,
    title = "Income Per Capita",
    position = "bottomright"
  )
```

The `addLayersControl()` function adds radio buttons to the map so readers can switch between years and observe change over time.

## Adding External Data Layers

`tidycensus` handles Census data beautifully, but often you'll want to overlay data from other sources — for example, historical boundaries from a city open data portal or academic dataset. Since `tidycensus` returns standard `sf` objects, you can layer any spatial data on the same map.

Here we map the [Home Owners' Loan Corporation (HOLC) redlining boundaries](https://dsl.richmond.edu/panorama/redlining/) for Los Angeles, from the [Mapping Inequality project](https://dsl.richmond.edu/panorama/redlining/) at the University of Richmond. These 1930s-era maps graded neighborhoods A–D, with "D" (red) areas systematically denied mortgage lending — a practice with lasting effects on wealth and segregation. The data is loaded directly from their public GeoJSON endpoint:

```{r, redlining}
me_la <- st_read(
  paste0(
    "https://dsl.richmond.edu/panorama/redlining/",
    "static/downloads/geojson/CALosAngeles1939.geojson"
  )
)

leaflet() |>
  addTiles(
    urlTemplate = paste0(
      "https://{s}.basemaps.cartocdn.com/",
      "light_all/{z}/{x}/{y}{r}.png"
    ),
    attribution = paste0(
      "&copy; OpenStreetMap contributors",
      " &copy; CARTO"
    )
  ) |>
  addPolygons(
    data = me_la,
    color = ~fill,
    fillColor = ~fill,
    fillOpacity = 0.6,
    weight = 1,
    label = ~paste("Grade:", grade)
  )
```

The `fill` column in this dataset already contains the HOLC grade colors (green for "A", red for "D"), so we pass it directly to both `color` and `fillColor`. This is an example of a self-describing dataset — the color encoding is built into the data.

You can then layer `get_acs()` data on top of this to compare current income or poverty rates against the historical redlining boundaries — a common and powerful technique in data journalism.

## Working With Non-Census GeoJSON

Many city and state open data portals publish data as GeoJSON that can be loaded directly into R via URL. For example, [Philadelphia's Neighborhood Food Retail dataset](https://opendataphilly.org/datasets/neighborhood-food-retail/) from [OpenDataPhilly](https://opendataphilly.org) includes supermarket access and restaurant counts by neighborhood:

```{r, food}
food <- st_read(
  paste0(
    "https://hub.arcgis.com/api/v3/datasets/",
    "53b8a1c653a74c92b2de23a5d7bf04a0_0/downloads/data",
    "?format=geojson&spatialRefId=4326&where=1%3D1"
  )
)

pal_restaurants <- colorNumeric(
  "Reds",
  domain = as.numeric(food$TOTAL_RESTAURANTS)
)
pal_supermarket <- colorFactor(
  palette = c("Yes" = "firebrick", "No" = "darkgreen"),
  domain = food$SUPERMARKET_ACCESS
)

leaflet(food) |>
  addTiles() |>
  addPolygons(
    data = food, group = "Restaurants",
    weight = 1, color = "black", fillOpacity = 0.8,
    fillColor = ~pal_restaurants(TOTAL_RESTAURANTS),
    label = ~paste0("Restaurants: ", TOTAL_RESTAURANTS)
  ) |>
  addPolygons(
    data = food, group = "Supermarkets",
    weight = 1, color = "black", fillOpacity = 0.8,
    fillColor = ~pal_supermarket(SUPERMARKET_ACCESS),
    label = ~paste0("Supermarket Access: ", SUPERMARKET_ACCESS)
  ) |>
  addLayersControl(
    baseGroups = c("Restaurants", "Supermarkets"),
    options = layersControlOptions(collapsed = FALSE)
  )
```

Notice that `colorFactor` is used here instead of `colorNumeric` or `colorBin` — because supermarket access is a categorical variable ("Yes"/"No"), not a number. Choosing the right palette function matters: use `colorNumeric` for continuous data, `colorBin` to group continuous data into ranges, and `colorFactor` for categorical data.

## Mapping Non-US Data

`tidycensus` is specific to U.S. Census data, but the same Leaflet workflow applies anywhere you can get spatial data. For example, local authority district boundaries for England are available as GeoJSON from [this GitHub repository](https://github.com/martinjc/UK-GeoJSON) and can be loaded directly from a URL:

```{r, england}
england <- st_read(
  paste0(
    "https://raw.githubusercontent.com/martinjc/",
    "UK-GeoJSON/refs/heads/master/json/",
    "administrative/eng/lad.json"
  )
)

england <- st_simplify(england, dTolerance = 100)
england <- st_transform(england, crs = 4326)

# attach whatever data you have
# here we use random values as a placeholder
england$data_value <- runif(nrow(england), 0, 100)

pal_england <- colorNumeric(
  palette = "YlOrRd",
  domain = england$data_value
)

leaflet(england) |>
  addTiles() |>
  addPolygons(
    fillColor = ~pal_england(data_value),
    fillOpacity = 0.7,
    color = "white",
    weight = 1,
    highlightOptions = highlightOptions(
      color = "black",
      weight = 2,
      bringToFront = TRUE
    ),
    label = ~paste0(LAD13NM, ": ", round(data_value, 1))
  ) |>
  addLegend(
    pal = pal_england,
    values = ~data_value,
    title = "Data Value"
  )
```

`st_simplify()` reduces the complexity of the polygon boundaries, which speeds up rendering significantly for large datasets. The `dTolerance` value controls how much simplification is applied — larger values are faster but less precise.

## Summary

The workflow for a tidycensus-based map follows these steps every time:

1. **Pull data** with `get_acs()`, setting `geometry = TRUE`
2. **Set the projection** with `st_transform('+proj=longlat +datum=WGS84')` so Leaflet can read it
3. **Define a color palette** with `colorNumeric`, `colorBin`, or `colorFactor` depending on your data type
4. **Build the map** with `leaflet()` + `addTiles()` + `addPolygons()`
5. **Add controls and legends** as needed with `addLayersControl()` and `addLegend()`

The advantage of this approach over manually downloading shapefiles is reproducibility: your code documents exactly which Census variables and years you used, and anyone running it will get the same data.
